---
title: "Reproducible Research - Assignment 1"
author: "Otavio Cals"
date: "November 13, 2015"
output: html_document
---

# Reproducible Research - Assignment 1
## Otavio Cals
### November 13, 2015

  
This is an R Markdown report for the first assignment of the Reproducible Research course. Here we will load a given database (activity.csv) and we will peform a simple data analysis on it. Each step of the analysis will be done in the order required by the assignment and explicitly shown.

# Part 1: Loading the Data

Here we will load the dataset and print it's top row. No data processing of the raw data has been deemed necessary.


```{r}
data <- read.csv("./activity.csv")
head(data,10)
```



# Part 2: Data Analysis by Day (NAs ignored)

In this part, we will perform a basic analysis of the variable "steps" recorded in the data by each day. Missing values (NAs) will be ignored.

We begin by calculating the total number of steps taken per day:

```{r}
sumstepsdate<-tapply(data$steps,data$date,sum,na.rm=TRUE)
print(sumstepsdate)
```

Now, we take the total number of steps taken per day data (sumstepsdate) and make a histogram out of it:

```{r}
hist(sumstepsdate,ylim=c(0,30),main="Histogram of Steps Taken in a Day",xlab="Total Number os Steps Taken in a Day")
```

Now, we calculate the mean and median of the total number of steps taken per day.

```{r}
meanstepsdate<-tapply(data$steps,data$date,mean,na.rm=TRUE)
medianstepsdate<-tapply(data$steps,data$date,median,na.rm=TRUE)
print(meanstepsdate)
print(medianstepsdate)
```


# Part 3: Data Analysis by Time Interval (NAs ignored)

In this part, we will perform a basic analysis of the variable "steps" recorded in the data by each time interval of 5 minutes. Missing values (NAs) will be ignored.

we begin by making a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):

```{r}
meanstepstime<-tapply(data$steps,data$interval,mean,na.rm=TRUE)
plot(names(meanstepstime),meanstepstime,main="Average Steps Taken",xlab="Time Interval",ylab="Average Steps",type = "l")
```

Now we calculate the period of time which,  on average across all the days in the dataset, contains the maximum number of steps:

```{r}
sumstepstime<-tapply(data$steps,data$interval,sum,na.rm=TRUE)
print(as.integer(names(sumstepstime[match(max(as.vector(sumstepstime)),as.vector(sumstepstime))])))
```

# Part 4: Inputing Missing Values

In this part, we will input specific values on the Missing Values (NAs) in the dataset.

We begin by counting the total number of NAs in the data:

```{r}
print(sum(is.na(data$steps)))
```

For this analysis, we decided that it would be more accurate to fill the missing data fields with the mean steps value of it`s corresponding time interval.
We begin by creating a copy of the data set and then input the missing values on it:

```{r}
data2<-data
for(i in 1:nrow(data2))
	{
		if(is.na(data2$steps[i]))
			{
				data2$steps[i]<-as.table(meanstepstime[toString(data2$interval[i])])
			}
}
head(data2,10)
```

We can verify that there is no longer any NAs in this new data by running the same code from before:

```{r}
print(sum(is.na(data2$steps)))
```

Now, we make a histogram of the total number of steps taken each day:

```{r}
sumstepsdate2<-tapply(data2$steps,data2$date,sum)
hist(sumstepsdate2,ylim=c(0,40),main="Histogram of Steps Taken in a Day",xlab="Total Number os Steps Taken in a Day")
```

Then we calculate the mean and median total number of steps taken per day:

```{r}
meanstepsdate2<-tapply(data2$steps,data2$date,mean)
medianstepsdate2<-tapply(data2$steps,data2$date,median)
print(meanstepsdate2)
print(medianstepsdate2)
```

We can thus verify that inputing data in the missing values made a impact in the new analysis performed.

# Part 5: Weekday/Weekend Steps Analysis

In this final part, we want to verify the differences in activity patterns between weekdays and weekends.

We will begin by creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day:

```{r}
weekdayslist<-weekdays(as.Date(data2$date),abbreviate=TRUE)
for(i in 1:length(weekdayslist))
	{
		if(weekdayslist[i] %in% c("Sun","Sat"))
			{
				weekdayslist[i]<-"weekend"
			}
		else
			{
				weekdayslist[i]<-"weekday"
			}
	}
data2<-cbind(data2,weekdayslist)
colnames(data2)[4]<-"weekdays"
head(data2,10)
```

Finally, we make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):

```{r}
meanstepstime2<-tapply(data2$steps,list(data2$interval,data2$weekdays),mean)
par(mfrow=c(1,2))
plot(as.integer(row.names(meanstepstime2)),meanstepstime2[,1],ylim = c(0,250),main="Average Steps on Weekdays",xlab="Time Interval",ylab="Average Steps",type = "l")
plot(as.integer(row.names(meanstepstime2)),meanstepstime2[,2],ylim = c(0,250),main="Average Steps on Weekends",xlab="Time Interval",ylab="Average Steps",type = "l")
```

We conclude then that the subject is more active during the weekends.
